An elementary and constructive proof of Grothendieck's generic freeness lemma

### 2018-04-17 paper-generic-freeness.tex
References

\title{An elementary and constructive proof of Grothendieck's generic freeness
lemma}
\author{Ingo Blechschmidt}

@@

@@

Previously known proofs proceed in a series of intermediate steps, reducing
to the case that~$A$ is a Noetherian integral domain where one can argue by
\emph{dévissage}; but in fact, a direct proof is possible and shorter.

@@

The new proof was found using a general topos-theoretical technique which we
believe to be useful in other situations as well. This technique allows to view
reduced rings and their modules from a different point of view, one from which
reduced rings look like fields. Since Grothendieck's generic freeness
is trivial for fields, this technique yields a trivial proof for reduced rings.
The proof presented here was obtained by unwinding the topos-theoretic proof,
yielding a self-contained argument without any references to topos theory.
We refer readers who want to learn about this technique to an accompanying
note~\cite{blechschmidt:wlog}.

\XXX{Acknowledgments, especially Martin Brandenburg}

\XXX{References to prior proofs: EGA, Stacks Project, Charles Staats. Check
\href{https://math.stackexchange.com/questions/1048496/question-on-generic-freeness-ref-matsumura-page-185}{this
MO question}}

@@

\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
\begin{align*}
  I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
  \quad\text{and} \\
  J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, \ell \in \{ 1,\ldots,m \}, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
\end{align*}
with the lexicographic order. The family~$(w_j)_{j \in J} \defeq (x_1^{i_1}
\ldots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ thus generates~$M$ as
an~$A$-module, and we'll call a subfamily~$(w_j)_{j \in J' \subseteq J}$
\emph{good} if and only if for all~$j \in J$, the vector~$w_j$ is a linear
combination of the vectors~$(w_{j'})_{j' \in J, j' \preceq j}$,
and if~$(\ell,i_1,\ldots,i_n) \not\in J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
I'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
how a good generating family can look like.

\begin{figure}
  \begin{tikzpicture}[scale=1.0]
    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

@@

    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
    \begin{scope}[yshift=4cm, xshift=4cm]
      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
      };
    \end{scope}
  \end{tikzpicture}
  \caption{\label{fig:good}A graphical depiction of a good generating family in
  the special case~$n = 2, m = 1$. The hatched cells indicate vectors which
  have already been removed from the family. If the vector in the red cell will
  be found to be expressible as a linear combination of vectors with smaller
  index (blue cells), it will be removed, along with the vectors in all cells
  to the top and to the right of the red cell. Corners are indicated by the
  small black squares.}
\end{figure}

@@

Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
generators with strictly smaller index. Then~$M$ is isomorphic to
the~$B$-module
\[ B\langle V_1,\ldots, V_m\rangle/
  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
\]
where the index~$j$ ranges over all corners of~$J$, and therefore finitely
presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
we mod out a suitable ideal with as many generators as corners of~$I$, is
isomorphic to~$B$.

@@

The following proposition is just an instance of Grothendieck's generic
freeness lemma. We prove it here because it admits an easier proof.

@@

>>>>

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-generic-freeness.bib}

\title[Grothendieck's generic freeness lemma]{An elementary and constructive
proof of Grothendieck's generic freeness lemma}
\author{Ingo Blechschmidt}

@@

\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

@@

Previously known proofs either only cover the case where~$A$ is a Noetherian
integral domain, where one can argue by \emph{dévissage} (see for
instance~\cite[Lemme~6.9.2]{ega-4-2},
\cite[Thm.~24.1]{matsumura:commutative-ring-theory}, or
\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
intermediate steps, reducing to that case (see for
instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
a direct proof is possible and shorter.

@@

The new proof was found using a general topos-theoretical technique which we
believe to be useful in other situations as well. This technique allows to view
reduced rings and their modules from a different point of view, one from which
reduced rings look like fields. Since Grothendieck's generic freeness
is trivial for fields, this technique yields a trivial proof for reduced rings.
The proof presented here was obtained by unwinding the topos-theoretic proof,
yielding a self-contained argument without any references to topos theory.
We refer readers who want to learn about this technique to a companion 
paper~\cite{blechschmidt:wlog}.

\textbf{Acknowledgments.} The proof presented here was prompted by
user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
benefited from discussions with Martin Brandenburg, who needed the constructive
version in paper of him~\cite{brandenburg:schur}. I'm grateful to Marc
Nieper-Wißkirchen for carefully guiding my PhD thesis, where \XXX{...}

@@

\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
\begin{align*}
  I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
  \quad\text{and} \\
  J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, \ell \in \{ 1,\ldots,m \}, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
\end{align*}
with the lexicographic order. The family~$(w_j)_{j \in J} \defeq (x_1^{i_1}
\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ generates~$M$ as
an~$A$-module, and we'll call a subfamily~$(w_j)_{j \in J' \subseteq J}$
\emph{good} if and only if for all~$j \in J$, the vector~$w_j$ is a linear
combination of the vectors~$(w_{j'})_{j' \in J, j' \preceq j}$,
and if~$(\ell,i_1,\ldots,i_n) \not\in J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
I'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
how a good generating family can look like.

\begin{figure}[b]
  \begin{tikzpicture}[scale=1.0]
    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

@@

    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
    \begin{scope}[yshift=4cm, xshift=4cm]
      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
      };
    \end{scope}
  \end{tikzpicture}
  \caption{\label{fig:good}A graphical depiction of a good generating family in
  the special case~$n = 2, m = 1$. The hatched cells indicate vectors which
  have already been removed from the family. The small black squares indicate
  \emph{corners}. If the vector in the red cell will be found to be expressible
  as a linear combination of vectors with smaller index (blue cells), it will
  be removed, along with the vectors in all cells to the top and to the right
  of the red cell.}
\end{figure}

@@

Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
generators with strictly smaller index. Then~$M$ is isomorphic to
the~$B$-module
\[ B\langle V_1,\ldots, V_m\rangle/
  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
\]
where the index~$j$ ranges over all corners of~$J$, and is therefore finitely
presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
we mod out a suitable ideal with as many generators as corners of~$I$, is
isomorphic to~$B$.

@@

The following proposition is just an instance of Grothendieck's generic
freeness lemma. We present it here because it admits an easier proof.

@@

\printbibliography
